---
title: "EV Power - Lab 4 Project Report"
format: typst
---

# Example Solution 1

## **Part 0: libraries**

```{r}
library(tidyverse)
library(janitor)
library(sf)
library(leaflet)
library(maps)
```



## **Part 1:** **Defining Research Question**

Chosen Question: Are electric vehicles more common in states where a larger share of electricity comes from renewable sources?* 

We’ll explore whether the proportion of renewable energy in each state correlates with the number of EV registrations in 2023.

## **Part 2: Data Preparation and Cleaning**

```{r}
# Show files
print(list.files("data", full.names = TRUE))
renew_2021_raw <- readr::read_csv("data/renew-use-2021.csv", show_col_types = FALSE) |> janitor::clean_names()
renew_2022_raw <- readr::read_csv("data/renew-use-2022.csv", show_col_types = FALSE) |> janitor::clean_names()
renew_2023_raw <- readr::read_csv("data/renew-use-2023.csv", show_col_types = FALSE) |> janitor::clean_names()

total_2021_raw <- readr::read_csv("data/total-use-2021.csv", show_col_types = FALSE) |> janitor::clean_names()
total_2022_raw <- readr::read_csv("data/total-use-2022.csv", show_col_types = FALSE) |> janitor::clean_names()
total_2023_raw <- readr::read_csv("data/total-use-2023.csv", show_col_types = FALSE) |> janitor::clean_names()

ev_2023 <- readr::read_csv("data/ev-registrations-by-state-2023.csv", show_col_types = FALSE) |> janitor::clean_names()
ev_2023 <- ev_2023 |>
  dplyr::rename(state = electric_vehicle_registrations_by_state_2023,
                ev_registrations = x2) |>
  dplyr::mutate(
    state = stringr::str_to_title(stringr::str_squish(state)),
    ev_registrations = suppressWarnings(as.numeric(gsub("[^0-9.]", "", ev_registrations)))
  )

prep_renew <- function(df) {

  val_col <- names(df)[stringr::str_detect(names(df), "(?i)renewable_use")]
  stopifnot(length(val_col) == 1)
  df |>
    dplyr::rename(state = state,
                  energy_source = energy_source,
                  renew_value = !!rlang::sym(val_col)) |>
    dplyr::mutate(
      state = stringr::str_to_title(stringr::str_squish(state)),
      renew_value = suppressWarnings(as.numeric(gsub("[^0-9.]", "", renew_value)))
    ) |>
    dplyr::select(state, energy_source, renew_value)
}

renew_2021 <- prep_renew(renew_2021_raw)
renew_2022 <- prep_renew(renew_2022_raw)
renew_2023 <- prep_renew(renew_2023_raw)

abbr_map <- tibble::tibble(state_abbrev = state.abb, state = state.name)

prep_total <- function(df) {
  df |>
    tidyr::pivot_longer(
      cols = -energy_source,
      names_to = "state_abbrev",
      values_to = "total_value"
    ) |>
    dplyr::mutate(
      state_abbrev = stringr::str_to_upper(stringr::str_squish(state_abbrev)),
      total_value = suppressWarnings(as.numeric(gsub("[^0-9.]", "", total_value)))
    ) |>
    dplyr::left_join(abbr_map, by = "state_abbrev") |>
    dplyr::filter(!is.na(state)) |>
    dplyr::mutate(state = stringr::str_to_title(state)) |>
    dplyr::select(state, energy_source, total_value)
}

total_2021 <- prep_total(total_2021_raw)
total_2022 <- prep_total(total_2022_raw)
total_2023 <- prep_total(total_2023_raw)

# Quick peek
head(renew_2023); head(total_2023); head(ev_2023)
```

## **Part 3: Joining / Pivoting Datasets for Analysis**

```{r}
# Sum across energy_source per state (2023 only)
renew_2023_sum <- renew_2023 |>
  dplyr::group_by(state) |>
  dplyr::summarise(renew_total = sum(renew_value, na.rm = TRUE), .groups = "drop")

total_2023_sum <- total_2023 |>
  dplyr::group_by(state) |>
  dplyr::summarise(total_energy = sum(total_value, na.rm = TRUE), .groups = "drop")

energy_2023 <- dplyr::inner_join(renew_2023_sum, total_2023_sum, by = "state") |>
  dplyr::mutate(renewable_share = 100 * renew_total / total_energy)

# Join EV counts
ev_energy_2023 <- energy_2023 |>
  dplyr::inner_join(ev_2023, by = "state") |>
  dplyr::mutate(state_key = tolower(stringr::str_squish(state)))

# Sanity check
nrow(ev_energy_2023); head(ev_energy_2023)
```

## **Part 4: Mapping Visualization**

```{r}
# Map polygons (ggplot2 helper)
us_map <- ggplot2::map_data("state") |>
  tibble::as_tibble() |>
  dplyr::mutate(state_key = region)

plot_data <- dplyr::left_join(us_map, ev_energy_2023, by = "state_key")

# Choropleth of renewable share
ggplot2::ggplot(plot_data, ggplot2::aes(long, lat, group = group, fill = renewable_share)) +
  ggplot2::geom_polygon(color = "white", linewidth = 0.2) +
  ggplot2::coord_quickmap() +
  ggplot2::scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90",
                               name = "Renewable Share (%)") +
  ggplot2::labs(title = "Renewable Electricity Share by State (2023)", x = NULL, y = NULL) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "right",
                 plot.title = ggplot2::element_text(face = "bold", size = 14))

# Scatter: EV registrations vs renewable share
scatter_df <- ev_energy_2023 |>
  dplyr::filter(!is.na(renewable_share), !is.na(ev_registrations))

ggplot2::ggplot(scatter_df, ggplot2::aes(x = renewable_share, y = ev_registrations)) +
  ggplot2::geom_point(color = "darkgreen", alpha = 0.75) +
  ggplot2::geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6) +
  ggplot2::labs(title = "EV Registrations vs Renewable Share (2023)",
                x = "Renewable Share (%)", y = "EV Registrations") +
  ggplot2::theme_minimal(base_size = 11)
```

## **Part 5: Analysis**

Despite multiple attempts to prepare, clean, and merge the datasets, the mapping and visualization steps consistently produced blank or incomplete diagrams. This outcome likely stems from formatting inconsistencies among the provided CSV files.

Although the visualizations could not be successfully rendered, the analytical intent was to explore whether states with higher renewable energy shares also exhibit greater electric vehicle adoption. The blank plots ultimately illustrate a key challenge in data analysis: reliable insights depend on consistent, well-structured data. 
